Temporal ordering and registration of images in studies of developmental dynamics

### 2015-01-30 image_analysis_paper_development2/SI_text.tex
updated paper

\subsection*{Image parameters}

The original image resolution is $1024 \times 1024$ pixels for the fixed images, and $512 \times 512$ pixels for the live images; all images are subsampled to $100 \times 100$ pixels for vector diffusion maps analysis.
%
A scale bar for a representative image is shown in \fig~\ref{fig:ap_dv}.
%
When imaging, the laser settings are calibrated so that images have good dynamic range, while not saturating the detector.
%
The same laser settings are maintained within each experiment.
%
The intensity of each channel is integer-valued, and ranges from $0$ (no signal) to $255$ (saturated). 
%
The Canny method \citep{canny1986computational} is used to detect the edges of the nuclear signal in each image (using the \texttt{edge} function in MATLAB\textsuperscript{\textregistered}), and images are cropped to these edges to remove any size effects.
%
A constant $10$-pixel-wide border is then added to each image. 
%
Contrast-limited adaptive histogram equalization (using the \texttt{adapthisteq} function with an $8 \times 8$ tile grid, and a uniform distribution for the intensities with a clip limit of 0.01) is used to normalize the intensities of the nuclear signal.
%
The nuclear signal is blurred using a disc filter with a $5$~pixel radius (using the \texttt{imfilter} function).
%
For the multichannel images, the nuclear signal is scaled by half (after normalization and blurring) relative to the other signals, so that the total nuclear signal across the entire image is of a comparable magnitude to the other signals.

@@

\begin{figure*}
\includegraphics{figS3}
\caption{Eigenvalue spectra. {\it (A)} Eigenvalue spectrum for the live imaging data set presented in \fig~3. Note that there is a gap after the third eigenvalue. {\it (B), (C)} Second and third embedding coordinates, respectively, versus the first embedding coordinate for the live imaging data set presented in \fig~3. Note that coordinates 2 and 3 are higher harmonics (and thus simple functions) of coordinate 1, and are therefore not informative about structure in the data set.  We can conclude that the data set is effectively one-dimensional. {\it (D)} Eigenvalue spectrum for the live imaging data set presented in \fig~4. Note that there is a gap after the first eigenvalue; we can conclude that the data set is effectively one-dimensional. }
\end{figure*}

\begin{figure*}
\includegraphics{figS4}
\caption{Live imaging data set. {\it (A)} Live imaging data set used for the results in \fig~3. The images have been preprocessed as described above, each image has been arbitrarily rotated, and the order of the images has been shuffled. {\it (B)} Data set from A after registration and ordering. The real time for each image is indicated. }
\end{figure*}

\begin{figure*}
\includegraphics{figS5}
\caption{{\it (A-D)} Four additional live imaging data sets used for the analysis in \fig~3. }
\end{figure*}

\begin{figure*}	
\includegraphics{figS6}
\caption{Fixed embryo imaging data set. {\it (A)} Fixed images used for the results in \fig~4. The images have been preprocessed as described above. {\it (B)} Data set from A after registration and ordering. The rank provided by a trained embryologist for each image is indicated. }
\end{figure*}

>>>>

\subsection*{Image preprocessing}

Before applying vector diffusion maps, images must be preprocessed so that the Euclidean distance between time image pixels is informative.
%
We want images who are developmentally similar to have relatively small Euclidean distances, and images who are at disparate developmental time points to have relatively large distances. 
%
We therefore must adjust images to remove any experimental and imaging artifacts, so that most of the variation is relevant to the developmental dynamics.

The relevant image operations are 
\begin{enumerate}
\item {\bf Intensity normalization:} Contrast-limited adaptive histogram equalization (using the \texttt{adapthisteq} function with an $8 \times 8$ tile grid, and a uniform distribution for the intensities with a clip limit of 0.01) is used to normalize the intensities of signals whose absolute intensity is not meaningful/informative.
%
\item {\bf Blur:} The \texttt{imfilter} function with a disc filter is used to blur signals whose small-scale structure is not informative. 
%
\item {\bf Intensity scaling (multichannel images only):} The \texttt{immultiply} function is used to scale signal intensities. This is important for multichannel images, and it determines the (relative) contributions of the various signals.  
%
\item {\bf Mean-centering:}  The Canny method \citep{canny1986computational} is used to detect the edges of the object in each image (using the \texttt{edge} function). The image is then translated so that the object (as determined by extremeties of the detected edges) is centered.
%
\item {\bf Size scaling:} For images whose relative size is unimportant to the developmental dynamics, the images are rescaled/dialated so that the object size (as determined by extremeties of the detected edges) is constant (we set this to be 80\% of the total image).
%
\item {\bf Removing corners:} The image is cropped/filtered using a disc centered in the middle of the image and whose diameter is equal to the number of pixels; any pixels outside of this disc are set to 0 (no intensity). This removes any corner effects under rotations. 
\end{enumerate}

\subsubsection*{{\em Drosophila} live imaging}

%
The original image resolution is $512 \times 512$ for the live {\em Drosophila} images.
%
All images were subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set. 
%
Images were normalized, and then blurred with a filter of radius 5\% of the total image (5 pixels). 
%
Images were not mean-centered or rescaled, as the entire live imaging data set was approximately centered already.

\subsubsection*{Zebrafish images}

The original image resolution was $320 \times 288$ for the zebrafish images.
%
All images were subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set. 
%
Images were not normalized or blurred.
%
Images were mean-centered so that the embryo in each frame was (approximately) centered. 
%
Images were not rescaled for size, as changes in overall were are important.

\subsubsection*{{\em Drosophila} fixed images}

The original image resolution was $1024 \times 1024$ for the fixed {\em Drosophila} images.
%
All images were subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set. 
%
The nuclear channel was normalized, and all channels blurred with a filter of radius 5\% of the total image (5 pixels) to remove the effects of individual nuclei. 
%
The nuclear channel was scaled by half; because this signal occupies a larger fraction of the image relative to the other signals, its overall contribution is large, and so we downscaled it so that each signal would have more comparable weight in the algorithm. 
%
The images were mean-centered and scaled to have a constant size using the nuclear signal to detect the edges of the embryo within the frame.

\subsubsection*{Wing disc images}

The original image resolution was $1024 \times 1024$ for the wing disc z-stacks.
%
Each of the original wing disc z-stacks contains 30--40 images. 
%
All images were subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set. 
%
Each wing disc z-stack was reduced to a 21-image stack, consisting of the brightest image and the 10 images above and below.
%
No channels were normalized or blurred in the images, and the channel intensities were kept at their imaging values. 
%
Images were mean-centered using the Wingless/Patched signal to detect the edges of the sample. 
%
Images were not rescaled for size, as changes in overall were are important.

%The original image resolution is $1024 \times 1024$ pixels for the fixed images, and $512 \times 512$ pixels for the live images; all images are subsampled to $100 \times 100$ pixels for vector diffusion maps analysis.
%%
%A scale bar for a representative image is shown in \fig~\ref{fig:ap_dv}.
%%
%When imaging, the laser settings are calibrated so that images have good dynamic range, while not saturating the detector.
%%
%The same laser settings are maintained within each experiment.
%%
%The intensity of each channel is integer-valued, and ranges from $0$ (no signal) to $255$ (saturated). 
%%
%The Canny method \citep{canny1986computational} is used to detect the edges of the nuclear signal in each image (using the \texttt{edge} function in MATLAB\textsuperscript{\textregistered}), and images are cropped to these edges to remove any size effects.
%%
%A constant $10$-pixel-wide border is then added to each image. 
%%
%Contrast-limited adaptive histogram equalization (using the \texttt{adapthisteq} function with an $8 \times 8$ tile grid, and a uniform distribution for the intensities with a clip limit of 0.01) is used to normalize the intensities of the nuclear signal.
%%
%The nuclear signal is blurred using a disc filter with a $5$~pixel radius (using the \texttt{imfilter} function).
%%
%For the multichannel images, the nuclear signal is scaled by half (after normalization and blurring) relative to the other signals, so that the total nuclear signal across the entire image is of a comparable magnitude to the other signals.

@@

\begin{figure}
\includegraphics[width=8cm]{drosophila_live_eval_spectrum}

\includegraphics[width=2.5cm]{drosophila_live_evec_corr1}
\includegraphics[width=2.5cm]{drosophila_live_evec_corr2}
\includegraphics[width=2.5cm]{drosophila_live_evec_corr3}
\caption{Eigenvalue spectra for the {\em Drosophila} live imaging data set presented in \fig~3. Note that there is a gap after the fourth eigenvalue product. Below are the second, third, and fourth embedding coordinate plotted versus the first embedding coordinate. Note that coordinates 2--4 are higher harmonics (and thus simple functions) of coordinate 1, and are therefore not informative about structure in the data set.  We can conclude that the data set is effectively one-dimensional and can be parameterized/ordered by coordinate 1.}
\end{figure}

\begin{figure}
\includegraphics[width=8cm]{zebrafish_eval_spectrum}

\includegraphics[width=2.5cm]{zebrafish_evec_corr1}
\includegraphics[width=2.5cm]{zebrafish_evec_corr2}
\includegraphics[width=2.5cm]{zebrafish_evec_corr3}
\caption{Eigenvalue spectra for the zebrafish data set presented in \fig~4. Note that there is a gap after the fourth eigenvalue product. Below are the second, third, and fourth embedding coordinate plotted versus the first embedding coordinate. Note that coordinates 2--4 are higher harmonics (and thus simple functions) of coordinate 1, and are therefore not informative about structure in the data set.  It is not immediately obvious that coordinate 3 is a harmonic of coordinate 1; the distortion in the plot is due to density effects in the data (the developmental changes are slower towards the beginning of the trajectory, and so there is a higher density of images in this portion of the one-dimensional curve), We can conclude that the data set is effectively one-dimensional and can be parameterized/ordered by coordinate 1.}
\end{figure}

\begin{figure}
\includegraphics[width=8cm]{drosophila_fixed_eval_spectrum}
\caption{Eigenvalue spectra for the fixed {\em Drosophila} images presented in \fig~5. Note that there is a gap after the first eigenvalue product. We can conclude that the data set is effectively one-dimensional and can be parameterized/ordered by coordinate 1.}
\end{figure}

\begin{figure}
\includegraphics[width=8cm]{wing_disc_eval_spectrum}
\caption{Eigenvalue spectra for the wing disc data set presented in \fig~6. Note that there is a gap after the first eigenvalue product. We can conclude that the data set is effectively one-dimensional and can be parameterized/ordered by coordinate 1.}
\end{figure}

%
%\begin{figure*}
%\includegraphics{figS3}
%\caption{Eigenvalue spectra. {\it (A)} Eigenvalue spectrum for the live imaging data set presented in \fig~3. Note that there is a gap after the third eigenvalue. {\it (B), (C)} Second and third embedding coordinates, respectively, versus the first embedding coordinate for the live imaging data set presented in \fig~3. Note that coordinates 2 and 3 are higher harmonics (and thus simple functions) of coordinate 1, and are therefore not informative about structure in the data set.  We can conclude that the data set is effectively one-dimensional. {\it (D)} Eigenvalue spectrum for the live imaging data set presented in \fig~4. Note that there is a gap after the first eigenvalue; we can conclude that the data set is effectively one-dimensional. }
%\end{figure*}
%
%\begin{figure*}
%\includegraphics{figS4}
%\caption{Live imaging data set. {\it (A)} Live imaging data set used for the results in \fig~3. The images have been preprocessed as described above, each image has been arbitrarily rotated, and the order of the images has been shuffled. {\it (B)} Data set from A after registration and ordering. The real time for each image is indicated. }
%\end{figure*}
%
%\begin{figure*}
%\includegraphics{figS5}
%\caption{{\it (A-D)} Four additional live imaging data sets used for the analysis in \fig~3. }
%\end{figure*}
%
%\begin{figure*}	
%\includegraphics{figS6}
%\caption{Fixed embryo imaging data set. {\it (A)} Fixed images used for the results in \fig~4. The images have been preprocessed as described above. {\it (B)} Data set from A after registration and ordering. The rank provided by a trained embryologist for each image is indicated. }
%\end{figure*}

### 2015-01-30 image_analysis_paper_development2/code_for_distribution/README/sample_input_file.tex
updated paper

\documentclass[12pt]{article}

\usepackage{amsmath, amsfonts, amssymb}
\usepackage{graphicx}

\usepackage{epstopdf}

@@

\section{Introduction}

This is code for the registration and ordering algorithms described in \\
``Temporal ordering and registration of images in studies of developmental dynamics,'' Carmeline~J.~Dsilva, Bomyi~Lim, Hang~Lu, Amit~Singer , Stanislav~Y.~Shvartsman, and Ioannis~G.~Kevrekidis

The code reads in a set of two-dimensional images or three-dimensional z-stacks, can perform a number of preprocessing operations, and then registers and/or orders the images using the algorithms described in the paper.

\section{Requirements}

All algorithms and analysis were implemented in MATLAB\textsuperscript{\textregistered} (R2013b, The MathWorks, Natick, Massachusetts).
%
The code requires MATLAB along with the Image Processing Toolbox.

\section{How To Use}

There are two ways to use the code
\begin{enumerate}
\item Interactive GUI.
\item Program a simple script.
\end{enumerate}

The basic steps are as follows:
\begin{enumerate}
   \item Read in images
   \item Preprocess the images (smooth and/or equalize channels, mean-center, etc.)
   \item Run vector diffusion maps to register and order preprocessed images
   \item (Optional) Save registered and ordered images
\end{enumerate}

\subsection{Interactive GUI}

\includegraphics[width=10cm, trim=0cm 2cm 5cm 0cm, clip]{gui_screenshot_initial.jpg}

There are five steps
%
\begin{enumerate}
\item Read Images
\item Preprocess Images
\item Analyze Images
\item Save Images
\item Average Trajectory
\end{enumerate}

\subsubsection{Read Images}

The fields are as follows
\begin{description}
\item[2D/3D] Select ``2D'' if your images are traditional 2D images. Select ``3D'' if your images are z-stacks.
%
\item[Image Directory] Enter the directory where your images or z-stacks are stored. This directory can be an absolute path (e.g. \texttt{C:/Documents/Images}) or a path relative to the directory where this code is stored (e.g. \texttt{../images})
%
\end{description}
\subsection{Script}

>>>>

% This LaTeX was auto-generated from an M-file by MATLAB.
% To make changes, update the M-file and republish this document.

\documentclass{article}
\usepackage{graphicx}
\usepackage{color}

\sloppy
\definecolor{lightgray}{gray}{0.5}
\setlength{\parindent}{0pt}

@@

\subsection*{Contents}

\begin{itemize}
\setlength{\itemsep}{-1ex}
   \item Basic Algorithm
   \item Read Images
   \item Show images
   \item Preprocess Images
   \item Calculate pairwise alignments
   \item Apply vector diffusion maps
\end{itemize}
\begin{verbatim}
clear all
close all
\end{verbatim}

\subsection*{Basic Algorithm}

\begin{par}
The basic steps are as follows:
\end{par} \vspace{1em}
\begin{enumerate}
\setlength{\itemsep}{-1ex}
   \item Read in images
   \item Preprocess the images (smooth and/or equalize channels, mean-center, etc.)
   \item Run vector diffusion maps to register and order preprocessed images
\end{enumerate}

### 2015-01-30 image_analysis_paper_development2/manuscript.tex
updated paper

\customlabel{fig:DV_view}{S1}
\customlabel{fig:1d_example}{S2}
\customlabel{fig:eigenvalues}{S3}
\customlabel{fig:full_movie_data}{S4}
\customlabel{fig:all_movies}{S5}
\customlabel{fig:full_fixed_data}{S6}

@@

\fig~\ref{fig:drosophila_live_imaging}A shows selected images from this live imaging data set, which contains $40$ consecutive frames taken at $30$~second time intervals at a fixed position within a single embryo (see \fig~\ref{fig:full_movie_data}).
%
Each image shows an optical cross-section near the posterior pole of a vertically oriented developing embryo, with the nuclei labeled by Histone-RFP.
%
Each frame was arbitrarily rotated, and the order of the frames was scrambled.
%
The task is now to register these images and order them in time to reconstruct the developmental trajectory.

@@

To assess the robustness of our proposed methodology, we repeated this procedure with four additional data sets (see \fig~\ref{fig:all_movies}) extracted from independent live imaging studies spanning the same developmental time period. 
%
The results are shown in \fig~\ref{fig:drosophila_live_imaging}E. 
%
The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than 95\%, indicating that our methodology can reproducibly order data of this type.

@@

We then applied our approach to a data set where the true rotational orientation and temporal order was not known {\it a priori}.
%
\fig~\ref{fig:drosophila_fixed_images}A shows selected images from a set of $120$ images of developing {\em Drosophila} embryos (see \fig~\ref{fig:full_fixed_data}) which cover a thirty minute time interval spanning late cellularization through gastrulation.
%
This data set is more complex than the live imaging data sets in that it contains significantly more images, each of which provides information about tissue morphology and the spatial distribution of two regulatory proteins.
%
Each image shows an optical cross-section of the posterior view of a {\em different} embryo at a different rotational orientation and fixed at a different (and unknown) developmental time.
%
The nuclei (gray) were labeled with DAPI, a DNA stain.
%
Embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor which specifies the cells of the future muscle tissue.
%
Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that, in this context, specifies a subset of neuronal cells \citep{Lim2013kinetics}.

\fig~\ref{fig:drosophila_fixed_images}B shows the selected images in \fig~\ref{fig:drosophila_fixed_images}A, now registered and ordered using vector diffusion maps \citep{singer2012vector}.
%
However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set (see \ref{fig:full_fixed_data}).
%
To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI). 
%
Each snapshot in the average trajectory is the (weighted) average of a group of successive images from the registered and ordered data set.
%
Sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.

@@

Up to this point, we have focused on two-dimensional imaging studies. 
%
However, it is not uncommon to encounter three-dimensional imaging data in the form of z-stacks. 
%
To illustrate how our methods can be extended to three-dimensional data, we analyzed a wing disc data set.

Although the theory of vector diffusion maps readily extends to three-dimensional data, calculation of the pairwise alignments rapidly becomes computational prohibitive for three dimensional data, as one must now search over three Euler angles, rather than a single rotation angle. 
%
However, ordering data via diffusion maps is readily possible for (registered) three-dimensional data. 
%
Therefore, we can apply our methodology to three-dimensional data by first registering the data using an alternative methodology (such as registering max projections, or aligning a single feature of the data), and then ordering the registered data using diffusion maps.

For the wing disc data, we first registered the three-dimensional data by computing the optimal rotations to align the maximum projections of the data using vector diffusion maps.
%
Because the maximum projections are two-dimensional images, this step is no more computationally intensive than the previous examples. 
%
We then used the computed alignments to register the full three-dimensional data, and then computed the diffusion maps embedding of the three-dimensional registered data. 
%
The results of ordering the data using the first (non-trivial) diffusion maps coordinate are shown in \fig~\ref{fig:wing_disc}.

The data was sorted into $6$ time cohorts by an expert; the time cohort for each image in \fig~\ref{fig:wing_disc}B are indicated. 
%
The rank correlation coefficient between the ordering from diffusion maps and the expert cohort sorting is 0.9427, indicating that our ordering is consistent with the expert. 
%
Furthermore, we can extract a stereotypical developmental trajectory by averaging; this trajectory is shown in \fig~\ref{fig:wing_disc}.

@@

The requisite user intervention and parameter tuning required for our method is relatively minimal.
%
As a first step, images must be preprocessed so that the Euclidean distance between the pixels is informative.
%
Our software provides several preprocessing options (such as blurring, rescaling, and mean-centering), as well as some guidance for what options to select depending on the system of interest. 
%
Two algorithmic parameters, the angular discretization to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{subfig:dmaps1} and \SI), must also be defined; we also provide some guidance on selecting these parameters.
%
We found that the results are robust to both of these parameters, and that consistent parameter values served well for all data sets presented here. 
%
Overall, the tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, and so this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.

@@

\subsection*{Experiments}
%
Oregon-R was used as wild type strains. 
%
Embryos were collected and fixed at 22$^\circ$C. 
%
Monoclonal rabbit anti-dpERK (1:100, Cell signaling) and rat anti-Twist (1:500, a gift from Eric Wieschaus) were used to stain proteins of interest. 
%
DAPI (1:10,000, Vector Laboratories) was used to visualize nuclei, and Alexa Fluors (1:500, Invitrogen) were used as secondary antibodies. 
%
Histone-RFP strain was used to obtain time-lapse movie of gastrulating embryos at 22$^\circ$C. 
%
Embryos were loaded to the microfluidic device with PBST to keep them oxidized.

\subsection*{Microscopy}
%
Nikon A1-RS scanning confocal microscope, and the Nikon 60x Plan-Apo oil objective was used to image embryos. 
%
Embryos were collected, stained, and imaged together under the same microscope setting. 
%
End-on imaging was performed by using the microfluidics device described previously \citep{chung2010microfluidic}.
%
Images were collected at the focal plane $\sim$90~$\mathrm{\mu m}$ from the posterior pole of an embryo (see \fig~\ref{fig:DV_view}).

\subsection*{Image preprocessing}
%
Image resolution was subsampled to $100 \times 100$ pixels for vector diffusion maps analysis.
%
Borders were cropped from the images using a Canny edge detector applied to the nuclear signal. 
%
Contrast-limited adaptive histogram equalization and a a 5-pixel radius disc filter were used to normalize and blur the nuclear signal.

@@

The authors thank Adam Finkelstein,  Thomas Funkhouser, and John Storey for helpful discussions.

>>>>

\customlabel{fig:DV_view}{S1}
\customlabel{fig:1d_example}{S2}
\customlabel{fig:eigenvalues_drosophila_live}{S3}
\customlabel{fig:eigenvalues_zebrafish}{S4}
\customlabel{fig:eigenvalues_drosophila_fixed}{S5}
\customlabel{fig:eigenvalues_wing_disc}{S6}

@@

\fig~\ref{fig:drosophila_live_imaging}A shows selected images from this live imaging data set, which contains $40$ consecutive frames taken at $30$~second time intervals at a fixed position within a single embryo.
%
Each image shows an optical cross-section near the posterior pole of a vertically oriented developing embryo, with the nuclei labeled by Histone-RFP.
%
Each frame was arbitrarily rotated, and the order of the frames was scrambled.
%
The task is now to register these images and order them in time to reconstruct the developmental trajectory.

@@

To assess the robustness of our proposed methodology, we repeated this procedure with four additional data sets extracted from independent live imaging studies spanning the same developmental time period. 
%
The results are shown in \fig~\ref{fig:drosophila_live_imaging}E. 
%
The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than 95\%, indicating that our methodology can reproducibly order data of this type.

@@

We then applied our approach to a data set where the true rotational orientation and temporal order was not known {\it a priori}.
%
\fig~\ref{fig:drosophila_fixed_images}A shows selected images from a set of $120$ images of developing {\em Drosophila} embryos which cover a thirty minute time interval spanning late cellularization through gastrulation.
%
This data set is more complex than the live imaging data sets in that it contains significantly more images, each of which provides information about tissue morphology and the spatial distribution of two regulatory proteins.
%
Each image shows an optical cross-section of the posterior view of a {\em different} embryo at a different rotational orientation and fixed at a different (and unknown) developmental time.
%
The nuclei (gray) were labeled with DAPI, a DNA stain.
%
Embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor which specifies the cells of the future muscle tissue.
%
Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that, in this context, specifies a subset of neuronal cells \citep{Lim2013kinetics}.

\fig~\ref{fig:drosophila_fixed_images}B shows the selected images in \fig~\ref{fig:drosophila_fixed_images}A, now registered and ordered using vector diffusion maps \citep{singer2012vector}.
%
However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set.
%
To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI). 
%
Each snapshot in the average trajectory is the (weighted) average of a group of successive images from the registered and ordered data set.
%
Sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.

@@

We have demonstrated that our algorithm can recover the underlying developmental dynamics in several different types of imaging data sets, including ones containing interembryo variability. 
%
Thus far, we have only looked at two-dimensional data sets where biological samples are imaged as a single confocal slice or as a phase contrast image. 
%
However, many developmental processes are inherently three-dimensional, and the z-axis often contains interesting dynamics which are independent of the dynamics in the x-y plane. 
%
Researchers often take multiple z-slices to visualize the three-dimensional structure of a sample;
with advances in imaging techniques, such three-dimensional data will be more common, and it is important to validtate that our algorithm can be applied to such data.

We used an existing three-dimensional data set of fixed {\em Drosophila} wing imaginal discs, which was published in \citet{hamaratoglu2011dpp}.  
%
Imaginal discs are groups of progenitor cells in fly larva that will transform into a specific organ, such as wings, legs, and eyes, in the adult fly during the pupal transformation. 
%
The wing disc is an imaginal disc that turns into a fly wing. 
%
The data set is composed of 46 fixed wing discs, whose developmental time range from 72 to 112~hours after fertilization. 
%
Each disc contains 21 z-slices taken at 1~$\mu$m intervals. 
%
The discs were stained with antibodies that recognize Spalt (red), Dad-GFP (green), and Wingless/Patched (gray) (\fig~\ref{fig:wing_disc}A). 
%
The disc grows in time; at long times, the disc becomes too large to fit completely within the frame and only a partial disc is imaged.

Although the theory of vector diffusion maps readily extends to three-dimensional data, calculation of the pairwise alignments rapidly becomes computational prohibitive for three dimensional data, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle). 
%
However, ordering data via diffusion maps is readily possible for (registered) three-dimensional data. 
%
Therefore, we can apply our methodology to three-dimensional data by first registering the data using an alternative methodology (such as registering maximum intensity projection, or aligning a single feature of the data), and then ordering the registered data using diffusion maps. 
%
Instead, we used maximum intensity projection and converted the image into 2D, and then used the angular synchronization on 2D data. 
%
For the wing disc data, we first registered the three-dimensional data by computing the optimal rotations to align the maximum intensity projections of the data using vector diffusion maps.
%
Because the maximum intensity projections are two-dimensional images, this step is no more computationally intensive than the previous examples. 
%
Other options for registering samples include using a specific localized signal (such as Twi in \fig~\ref{fig:drosophila_fixed_images}) as a marker, or using other {\em a priori} knowledge.
%
We then used diffusion maps to order the registered three-dimensional images. 
%
\fig~\ref{fig:wing_disc}B shows selected images from the data set ordered by diffusion maps. 
%
In the original data set, each disc was assigned to 6 time classes (72--73~hr, 76.5--77.5~hr, 79--80~hr, 89--90~hr, 100--101~hr, and 110.5--111.5~hr after fertilization); these times are indicated in \fig~\ref{fig:wing_disc}B. 
%
In the ordered set, the size of wing disc grows, and the intensity of proteins increases as a function of time. 
%
The rank correlation coefficient based on the time class is 0.9427. 
%
Furthermore, we can extract a stereotypical developmental trajectory by averaging; this trajectory is shown in \fig~\ref{fig:wing_disc}.
%
Thus, this result shows that the algorithm can successfully order a three-dimensional data set.

@@

The requisite user intervention and parameter tuning required for our method is relatively minimal.
%
As a first step, images must be preprocessed so that the Euclidean distance between the pixels is informative.
%
Our software provides several preprocessing options (such as blurring, rescaling, and mean-centering), as well as some guidance for what options to select depending on the system of interest. 
%
Two algorithmic parameters, the angular discretization to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{fig:schematics}A and \SI), must also be defined; we also provide some guidance on selecting these parameters.
%
We found that the results are robust to both of these parameters, and that consistent parameter values served well for all data sets presented here. 
%
Overall, the tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, and so this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.

@@

\subsection*{{\em Drosophila} Experiments}
%
Oregon-R was used as wild type {\em Drosophila} strains. 
%
Embryos were collected and fixed at 22$^\circ$C. 
%
Monoclonal rabbit anti-dpERK (1:100, Cell signaling) and rat anti-Twist (1:500, a gift from Eric Wieschaus) were used to stain proteins of interest. 
%
DAPI (1:10,000, Vector Laboratories) was used to visualize nuclei, and Alexa Fluors (1:500, Invitrogen) were used as secondary antibodies. 
%
Histone-RFP strain was used to obtain time-lapse movie of gastrulating embryos at 22$^\circ$C. 
%
Embryos were loaded to the microfluidic device with PBST to keep them oxidized.

\subsection*{{\em Drosophila} Microscopy}
%
Nikon A1-RS scanning confocal microscope, and the Nikon 60x Plan-Apo oil objective was used to image embryos. 
%
Embryos were collected, stained, and imaged together under the same microscope setting. 
%
End-on imaging was performed by using the microfluidics device described previously \citep{chung2010microfluidic}.
%
Images were collected at the focal plane $\sim$90~$\mathrm{\mu m}$ from the posterior pole of an embryo (see \fig~\ref{fig:DV_view}).

\subsection*{Image preprocessing}

Images were subsampled, normalized, blurred, and centered prior to diffusion maps analysis to remove any variations due to the epxperimental and imaging framework. 
%
Details about the preprocessing operations applied to each imging data set are given in \SI.

@@

The authors thank Fisun Hamaratoglu and Markus Affolter for providing the wing disc data. 
%
The authors thank Angela DePace, Granton Jindal, Adam Finkelstein,  Thomas Funkhouser, and John Storey for helpful discussions.
